Thermal reactionomes reveal adaptive responses to thermal

### 2014-10-03 doc/ApTranscriptome_manuscript.Rmd
updates to Intro from SHC

---
title: Investigation of thermal reactionomes reveals differing responses to thermal
  extremes in warm and cold-climate ant species
author: ' '
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change. Here, we describe the thermal reactionome, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient, to identify genes underlying thermal tolerance. We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change than did *A. picea* in response to warming. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *Aphaenogaster carolinensis* also shut down more genes in response to thermal extremes than did *A. picea*, while a subset of these same genes were upregulated in *A. picea*, including gene ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences indicate that the two closely related species employ different strategies for coping with temperature stress. The northern *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature (sunday, kingsolve, deutch, hoffman). Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...). Thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of such variation or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. To address this gap of knowlege, we need information on the diversity of genes, and their plasticity of expression, involved in thermal tolerance [@hoffman2008; @somero2011].

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, many patterns of biologically-relevant plasticity are likely to be missed in studies that compare only a few levels of an environmental variable [@murren2014]. Similarly, species comparisons look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. Such differences in degree are better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster picea* are found ... while *A. carolinensis* occupy habitats with ... . Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To gain insight on the evolution of thermal adaptation, we compared the thermal reactionomes of these species to test two specific predictions. First, we predicted that species from more thermally-stressful environments will rely to a greater extent than species from cooler environments upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support this prediction would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

While neither of these predictions depend on the functional annotation of the responsive genes [@pavlidis2012], we used gene set enrichment analysis to gain insight on the molecular toolkit involved in adaptation to thermal extremes. For example, in response to thermal extremes, we expect genes in the cellular stress response [@kultz2005] to be up-regulated, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated [@somero2011].

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-26.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
```

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at 5% False Discovery Rate. Of these, `prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. This result is the same if we focus on only the transcripts that have a BLAST hit. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts (i.e. the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]) indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

@@

We found that the molecular diversity of thermal response to cold temperatures is substantially greater than of that to hot temperatures. In particular, we grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). In both species,  the majority of genes were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).

@@

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A. picea* but not *A. carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).

@@

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Though we found `r resp.stress.responsive.count` genes annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r GO.stress.Xsq$statistic`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012]. In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014] which only used a few temperature conditions. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

Our results are at congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher basal level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. There are 3-4 $\times$ more genes upregulated at cold than high temperatures in both species (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, these limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to expectations that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both species. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

Thermal safety margins...

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.

@@

...ants...Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012]. Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013].

@@

From each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013]. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

to each gene. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005]. This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.

We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal
  extremes in warm and cold-climate ant species
author: ' '
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change. To understand how plasticity of gene expression underlies thermal tolerance, we describe the *thermal reactionome*, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient. We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher mean levels of expression for responsive genes, exhibited greater inertia of gene expression change and upregulated fewer heat shock proteins than did *A. picea* in response to thermal extremes. This pattern is consistent with hypotheses of constitutive thermal defense and enhanced proteome stability. *Aphaenogaster carolinensis* also down-regulated expression of more genes in response to thermal extremes than did *A. picea*. A subset of these same genes were upregulated in *A. picea*, including genes mapped to ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences indicate that the two closely related species employ different strategies for coping with temperature stress. The northern *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...). Thermal defense strategies are shaped by the environmental regime organisms experience (...ref...), and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].

To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. hot vs. cold) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A. picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...). Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to test two specific predictions. First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms relative to species from cooler environments. [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_2014-09-26.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

options(digits = 2)
panderOptions('round', 2)
panderOptions('keep.trailing.zeros', TRUE)
```

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts (hereafter referred to as genes) with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. In effect, `r round(9100/99861, 2)*100`% of the transcriptome is thermally-responsive. This result is the same if we focus on only the transcripts that have a BLAST hit. However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] indicating that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

@@

The number of genes that were responsive to cold temperatures was substantially greater than of the number responsive to hot temperatures. We grouped the thermally-responsive genes for each species into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). In both species,  the majority of genes were in the **Low** category (Table 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Table 1).

@@

Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. While some of these genes were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` in *A. picea* but not *A. carolinensis*), most of these were genes that changed from one expression category to another between the species (i.e. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1).

@@

We examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` genes were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012]. Finding that thousands of genes responded to temperature highlights the limitations of previous studies [@meyer2011; @barshis2013; @palumbi2014; @oneil2014] that used only a few temperature levels and found that only a few dozen to hundred genes are involved in thermal responses. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher basal level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

Given the differences in the number of thermally-responsive genes between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more genes upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal gene expression in *A. picea* for genes that were up-regulated in response to cold temperatures (**Low** genes) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** genes in *A. picea* as well as for genes that were **Intermediate** in both species. A potential explanation for the consistently higher basal gene expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater gene expression.

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may have only small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.

An additional concern is differences in gene expression among species due to drift. While we do in fact expect some genes to differ in expession due to neutral processes, if this was the primary driver of differences among species we would expect to see equal distribution of counts in the off-diagonals of the matched observations (Fig. 1). Instead we find that certain categories are over-represented, and that the differences in the reaction norm shapes are consistent with *a priori* expectations.

@@

Ants of the genus *Aphaenogaster* are some of the most abundant in eastern North America [@lubertazzi2012], and species as well as populations within species differ in the critical maximum and minimum temperatures [@warren2013]. Temperature is a potentially strong selective force for ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].

@@

From each species, we haphazardly collected 12 ants from a single colony at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath maintained at one of 12 randomly-assigned temperatures (0° to 38.5°C, in 3.5° increments) for one hour. The minimum and maximum temperatures were selected based on previous work showing that these temperatures are close to the critical minimum and maximum temperatures for *Aphaenogaster* [@warren2013]. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. We extracted mRNA by homogenizing three pooled ants in 500 uL of RNAzol buffer with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY), followed by RNAzol extraction (Molecular Research Center Inc; Cincinnati, OH) and then an RNeasy micro extraction (Qiagen Inc; Valencia, CA) following the manufacturer's instructions.

@@

to each gene. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects [@cottingham2005]. This method is robust to over-dispersion because we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.

We identified thermally-responsive genes as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
